A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-25 paper/bare_conf.tex
Re-factor lots of paragraphs

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method and Finite Element Method, are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell of the discretized domain by applying suitable numerical kernels. Iteration over mesh components is also expensive, due to the presence of indirect memory accesses. As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this issue, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.

OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (``kernels''). A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Together with the indirections problem, kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in Finite Element Methods.

We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method). With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.

Since loop nests are affine, one possibility is to rely on polyhedral compilation for improving performance. There are, however, several limitations. The first problem is that a polyhedral compiler reasons on iteration spaces, whereas some of the transformations we propose rely on analysis of the data space. Our attempts at using PoCC's optimizations~\cite{PoCC}, especially register tiling, were unsuccessful. A domain-specific transformation of the iteration space, which is not expressible in the polyhedral model because of its non-affine nature, also showed a significant performance gain due to improved register locality.

Assembly kernels could be turned into a sequence of calls to BLAS routines, although it is complicated to achieve and, potentially, ineffective. One issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern. In addition, register locality would be affected, since the same matrix can appear in a number of products. Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}. Hand-made implementations of the simple Helmholtz assembly kernel as a sequence of BLAS $dgemm$ calls, for a range of polynomial orders ($p=1,2,3,4$), has showed that the achieved performance is worse than that exposed in Section~\ref{sec:perf-results}.

Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems. We have developed a domain-specific optimizing compiler and we have integrated it with Firedrake, a system for solving PDEs through the Finite Element Method~\cite{firedrake} based on the OP2 abstraction. This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).

@@

Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain. Numerical integration algorithms are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}.

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. At the core of Firedrake~\cite{firedrake} lies PyOP2~\cite{pyop2ics}, which triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh. In our work, we have enhanced this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used. Notable examples are~\cite{uflacs, fluidity, mcfc}, in which the kernel's loop nest and memory access pattern are reducible to that generated by Firedrake.

The complexity of Firedrake-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might have bended elements that require the Jacobian be re-computed every $i$ iteration. 2) The stiffness matrix $A$ is always the result of an outer-product computation along the $j$ and $k$ dimensions. 3) Memory accesses along the three loop dimensions are always 1-stride .4) The $j$ and $k$ loops are interchangable, whereas to permute $i$ it might be necessary to precompute $F$ values and to introduce temporaries. 5) Basis functions arrays (denoted by $X$, $Y$, ...) are constants, and many sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

% remember: op vectorization PLUS unroll and jam !!
\subsection{Outer-product Vectorization}
Register tiles along both inner dimensions can be determined as well. Our domain-driven strategy, based on the observation that the evaluation of the stiffness matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, aims at optimizing register reuse, in terms of both minimization of load operations and achievement of larger tiles, over a canonical implementation. If we consider the code snippet in Figure~\ref{code:helmholtz-licm}, we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of other constant or $i$-dependent variables can be momentarily neglected. Note that regardless of the specific input problem, the assembly computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix with just 2 vector registers. Once the first four entries are evaluated, values in a register are shuffled and the same product is performed again. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, would employ 5 registers, 4 of them target of a broadcast operation (i.e. ``splatting'' a single value over the four locations of an AVX register) along the outer dimension, and 1 of a load operation along the inner dimension. In our outer-product vectorization ($op$-$vect$), the 4 broadcast accesses to the L1 cache are avoided by an initial load and by shuffling. More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e. all $x$ terms in Equation~\ref{outer-product}), with $vl$ being the vector length. This is a considerable gain, which allows us to slice the iteration space into bigger tiles, implemented directly by vector registers.

>>>>

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method and Finite Element Method, are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell of the discretized domain by applying suitable numerical kernels. As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.

OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (``kernels''). A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted by~\cite{fenics, fluidity, lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in Finite Element Methods.

We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method). With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation. Our study is conducted in the context of Firedrake, a system for solving PDEs through the Finite Element Method based on the OP2 abstraction~\cite{firedrake}.

Optimization of assembly kernels is non-trivial. Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only by an in-depth analysis of the data space, rather than reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, were unsuccessful). BLAS routines could be employed as well, although fairly complicated code manipulations would be required, for a performance gain potentially far from the expectations. The major issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern. Register locality would also be affected, since the same matrix could appear in a number of products. BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}. Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$, has showed that the performance is worse than that achieved with our optimization strategy.

%Also, tiling at the level of vector register, achievable through a non-affine transformation of the iteration space, has showed a significant performance improvement in a range of problems. .

Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems. We have developed an optimizing compiler and we have integrated it with Firedrake. This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).

@@

Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain. Numerical integration algorithms are widely-used in assembly codes to implement evaluation of stiffness matrices~\cite{quadrature1, quadrature2}.

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh. In our work, we have enhanced this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used. Notable examples are~\cite{uflacs, fluidity, mcfc}, in which the kernel's loop nest and memory access pattern are reducible to that generated by Firedrake.

The complexity of Firedrake-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. The stiffness matrix in the code is called $A$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, and so infinite is the space of codes that Firedrake can generate, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might use bended elements that would require the Jacobian be re-computed every $i$ iteration. 2) Memory accesses along the three loop dimensions are always 1-stride .3) The $j$ and $k$ loops are interchangeable, whereas permutation of $i$ is might be subjected to pre-computation of values and introduction of temporaries (e.g. the $F$ values in Burgers). 4) So called basis functions arrays (denoted by $X$, $Y$, ...) are constants, and most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

% remember: op vectorization PLUS unroll and jam !!
\subsection{Outer-product Vectorization}
Register tiles along both inner dimensions can be determined as well. Based on the observation that the evaluation of the stiffness matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, aims at optimizing register reuse, in terms of both minimization of load operations and achievement of larger tiles, over a canonical implementation. If we consider the code snippet in Figure~\ref{code:helmholtz-licm}, we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of other constant or $i$-dependent variables can be momentarily neglected. Note that regardless of the specific input problem, the assembly computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix with just 2 vector registers. Once the first four entries are evaluated, values in a register are shuffled and the same product is performed again. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, would employ 5 registers, 4 of them target of a broadcast operation (i.e. ``splatting'' a single value over the four locations of an AVX register) along the outer dimension, and 1 of a load operation along the inner dimension. In our outer-product vectorization ($op$-$vect$), the 4 broadcast accesses to the L1 cache are avoided by an initial load and by shuffling. More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e. all $x$ terms in Equation~\ref{outer-product}), with $vl$ being the vector length. This is a considerable gain, which allows us to slice the iteration space into bigger tiles, implemented directly by vector registers.
